---
title: "IST 5535 Lab 1: College Data"
author: "Sayantan Majumdar"
date: "`r Sys.Date()`"
output: 
  html_document:
    number_sections: true
---

# Read Data
```{r}
college <- read.csv('College.csv')
typeof(college)
knitr::kable(head(college))
```

# Organizing Data
```{r}
rownames(college) <- college[,1]
college$X <- NULL
knitr::kable(head(college))
```

# Data Exploration

## Summary Statistics

```{r}
summary(college)
```

## Scatterplot Matrix

```{r}
pairs(college[,1:10])
```

## Boxplot
```{r}
outstate <- college$Outstate
private <- college$Private
boxplot(outstate~private, xlab='Private', ylab='Outstate', main='Outstate vs Private Boxplot')

```

## Elite Students
```{r}
univ.types <- ifelse(college$Top10perc > 50, "Elite", "Non-Elite")
Elite <- univ.types[univ.types == "Elite"]
summary(Elite)
```
Number of elite universities = `r length(Elite)`

## Histogram Plots
```{r}
par(mfrow=c(2, 2))
hist(college$Enroll, main="Enroll Histogram")
hist(outstate, main="Outstate Histogram")
hist(college$Grad.Rate, main="Grad.Rate Histogram")
hist(college$Books, main="Books Histogram")
```

## Using ggplot2
```{r}
library(ggplot2)

par(mfrow=c(1, 2))
g1 <- ggplot(college, aes(x=Grad.Rate, y=Room.Board, col=univ.types)) + geom_point() + geom_smooth(method="lm", se=F)
plot(g1)
```

The graduation rate is typically higher with medium room and board costs (between \$3500 and \$5000) as observed from the central cluster. However, strong non-linearity is present.

``` {r}
g2 <- ggplot(college, aes(x=Grad.Rate, y=univ.types)) + geom_point()
plot(g2)
```

Since the number of non-elite universities are higher so we see a greater number of points in the non-elite class. Moreover, the elite universities have higher graduation rates typically between 60-90% whereas for non-elite universities graduation rates vary a lot. So the earlier relationship with room and board costs has changed significantly.

```{r}
g3 <- ggplot(college, aes(x=Grad.Rate, y=Top10perc)) + geom_point()
plot(g3)
```

Interestingly, the graduate rate follows a strong quadratic trend when compared with the new students from top 10% of high school class which includes both elite and non-elite university students.
